---
title: "analysis"
output: html_notebook
---

We will do an analysis of the dataset first, and next we will answer some predefined questions.

Import data:

```{r}
full <- readxl::read_xlsx('covidrj.xlsx', progress = FALSE)
mortes <- diff(full$deaths)
plot(mortes, main = 'Deaths')
```

# Trend

We will now model $Z(t) = T(t) + a_t$.

##  Linear model

```{r}
t = c(1:length(mortes))
deaths = data.frame(mortes = mortes, t = t)
mod = lm(mortes ~ t, data=deaths)
summary(mod)
plot(deaths$t, deaths$mortes, main = 'Linear model')
abline(mod$coefficients)
plot(mod$residuals, main = 'Residuals of the linear model')
```

It seems not a good a idea.

## Polynomial model

```{r}
mod = lm(mortes ~ poly(t, 2), data = deaths)
summary(mod)
prediction = predict(mod, data.frame(deaths$t))
plot(deaths$mortes, main = 'Polynomial model')
lines(prediction)
plot(mod$residuals, main = 'Residuals of the polynomial model')
```

We see better results.

## Moving average

```{r}
w1 = filter(x = deaths$mortes, filter = rep(1/7, 7), sides = 2)
w2 = filter(x = deaths$mortes, filter = rep(1/14, 14), sides = 2)
w3 = filter(x = deaths$mortes, filter = rep(1/39, 30), sides = 2)
plot(deaths$mortes, main = 'Moving average')
lines(w1, col='red')
lines(w2, col='blue')
lines(w3, col='green')
legend(
  x = 0,
  y = 300,
  legend = c('Data', 'w1', 'w2', 'w3'), 
  col = c('black', 'red', 'blue', 'green'),
  pch = c('o', '', '', ''),
  lty = c(0, 1, 1, 1)
)
```

## LOWESS

```{r}
lw1 = lowess(x = deaths$mortes, f=0.5)
lw2 = lowess(x = deaths$mortes, f=0.2)
plot(deaths$mortes, main = 'LOWESS')
lines(lw1, col='red')
lines(lw2, col='blue')
legend(
  x = 0,
  y = 300,
  legend = c('Data', 'lw1', 'lw2'), 
  col = c('black', 'red', 'blue'),
  pch = c('o', '', ''),
  lty = c(0, 1, 1)
)
```

## Differentiation and ACF

```{r}
dz = diff(deaths$mortes, lag = 1)
plot(dz, main = 'Differentiation')
acf(x = dz)
acf(deaths$mortes)
```

Diff is a good idea.

## Trend test

```{r}
library(randtests)
runs.test(x = deaths$mortes, alternative = "two.sided", plot = TRUE)
```

# Seasonality

We will now model $Z(t) = T(t) + S(t) + a_t$.

## Deterministic seasonality

```{r}
# Seasonal dummies
N = length(deaths$mortes)
Q = factor(c(rep(c(1:7), N/7), c(1:(N%%7))))
deaths$Q = Q
mod = lm(mortes ~ t+Q, data = deaths)
summary(mod)
plot(deaths$mortes, main = "Seasonality")
lines(mod$fitted.values)
```

## Log scale (multiplicative)

```{r}
index = deaths$mortes != 0
mod = lm(log(mortes[index]) ~ t[index]+Q[index], data = deaths)
summary(mod)
plot(deaths$mortes, main = "Seasonality (multiplicative)")
lines(exp(mod$fitted.values))
```

Better in logarithmic scale.

## Stochastic seasonality

```{r}
trend = filter(deaths$mortes, sides = 2, filter = rep(1/7, 7))
mod = lm(mortes ~ Q-1, data=deaths)
summary(mod)
S = mod$fitted.values
plot(deaths$mortes)
plot(trend)
plot(S)
plot(deaths$mortes - trend - S)
```

## Decompose

```{r}
plot(decompose(ts(deaths$mortes, frequency = 7)))
```

## LOWESS

```{r}
plot(stl(ts(deaths$mortes, frequency = 7), s.window = 7))
```

## Kruskal-Wallis

```{r}
trend = decompose(ts(deaths$mortes, frequency = 7))$trend
deaths$detrended = deaths$mortes - trend
kruskal.test(detrended ~ Q, data = deaths)
```

We reject the hypothesis of non seasonality.

## Analysis of variance

```{r}
mod <- lm(detrended~Q-1, na.action = na.omit, data = deaths)
summary(mod)
```

## Questions

1. Is the series stationary?

Clearly we see we have trend and seasonality, so it's not stationary.

2. Is there some seasonality?

Sure! All seasonality models show some pattern. Some of them have coefficients significative (greater than 0 with high p-value). We also did a Kruskal-Wallis and rejected the null hypothesis of non stationarity.

3. Is there some trend?

There is a trend. All models shows that. We also did a test and confirmed that.

4. Is logarithmic scale model better (multiplicative model)?

The $R^2$ for log scale is better.

5. Does the sequence test show trend?

Yes.

6. Do any of the tests studied in class show seasonality?

Yes.